Optimal Symmetric Multimodal Templates and Concatenated Random Forests for Supervised Brain Tumor Segmentation (Simplified) with \textit{ANTsR}

### 2014-04-29 paper.tex
ENH: More edits.

\begin{figure}
  \centering
  \begin{tabular}{c}
%    \includegraphics[width=88mm]{pipeline2.pdf}
    \includegraphics[width=110mm]{pipeline2.pdf}
  \end{tabular}
  \caption{Diagrammatic workflow for the proposed RF model training and prediction.  The feature
  images are first generated from the input MRI and symmetric multivariate template. If training,
  the set of feature images derived from the training data are used to create the RF model
  for the first stage.  The model is then applied to the training or prediction data to generate 
  a second set of feature images to create the second stage RF model (if training) and/or generate
  the second RF model.  
  For prediction, the final refinement process entails a series of heuristically-derived
  binary morphological resulting in the final labeled brain.
  \textcolor{red}{I still find this unclear, especially the arrows
    that cross between training and testing --- intuitively, this
    seems impossible w/o circularity.}
  }
  \label{fig:pipeline}
\end{figure}

We use random forests as supervised learners.  Given ground truth,
voxel-wise $n$-class labels for a specific image, we can vectorize the
label image to produce $v_L$, the $p$-length vector of class labels
where $p$ is the number of voxels in the given brain of interest.
We also assume a matrix of features for the given brain, which we
denote $\bm{F}$ where $\bm{F}$ has dimensionality of $p$ by
$k$, where each of the $k$ columns corresponds a potentially valuable
feature vector.  Given this representation, the random forest model
can be trained, in \textit{R} notation, to predict the labels via
$v_L \sim  \bm{F}$.
Our segmentation protocol involves training and application
of two RF models in succession 
(see Figure \ref{fig:pipeline}) to which we refer as ``stages.''  The basic idea is that
we generate a set of feature images used as input to the
first RF model (or first stage) which produces a voxelwise 
probabilistic tissue estimate.  More precisely, in the RF framework, each prediction sample (i.e. the feature vector at each voxel),
is propagated through each tree of the ensemble where it is labeled as belonging to a specific
class.  
These ``votes'' are converted to voxelwise probabilistic
estimates for each class via standard mechanisms used in random forest
models, e.g. \cite{liaw2002}.
The seven labels of brain/tumor tissue type denote:
\begin{itemize}
\item cerebrospinal fluid,
\item grey matter,
\item white matter,
\item edema, 
\item non-enhancing tumor (including low-grade tumor center), 
\item enhancing tumor (excluding necrotic center), and 
\item abnormal necrotic center or necrocyst in high-grade gliomas.
\end{itemize}
We then use these output tissue map estimates as spatial 
priors for generating a second set of geometrically refined image features.  These
are used as input for a second RF model application (or
second stage),
the output of which constitute the final tumor segmentation estimate.
Note that some feature images are included \textbf{directly and
  without modification(?)} in both stages such as
the asymmetry features described in the previous section.

@@

\paragraph{Stage 1 voxelwise classification with Gaussian mixture
  modeling} 
The first step in processing a new image is to segment the image into
the $n$ tissue classes listed above.   We achieve this via an expectation-maximization
segmentation method based on
Gaussian mixture modeling (GMM).  GMM results in voxelwise
tissue probabilities which comprise an additional set of image
features.  The GMM is initialized with prior cluster centers for
specific tissue types.  We learn the values for these cluster centers
for each modality and each tissue from training data
\citep{reynolds2009}.  We also perform intensity normalization
across the training cohort by winsorizing the intensity values to the quantile
range $[0.01, 0.99]$ and then rescaling the resulting intensities to 
$[0,1]$.  The cluster centers are defined as the mean
normalized intensity value for each tissue type of each modality 
image over all the training data.  It should be noted that 
initializing with these values is optional and that results
seem to be robust to initialization values.  \textcolor{red}{How else
  would you initialize? how robust?  how did you test this?}

More formally, the GMM computes the 
probability distribution at each voxel, $\mathbf{x}$, as the
sum of $M$ Gaussian components, $\mathcal{N}(\mathbf{x}|\mu,\sigma)$, i.e.
\begin{align}
p\left(\mathbf{x}|\mu_m,\sigma_m,\lambda_m\right) = \sum_{i=1}^M \lambda_m \mathcal{N}(\mathbf{x}|\mu_m,\sigma_m)
\end{align}
where $\sum_{m=1}^M \lambda_m = 1$.  The parameters of the GMM 
are determined using the 
Atropos segmentation tool \citep{avants2011} available in ANTs.
\textcolor{red}{What modalities are input to Atropos?  Should we
  reference a figure here to make this clear?  Would listing a table of the
  prior per-class per-modality tissue cluster centers be helpful?}

\paragraph{Stage 1 random forest} 
The stage 1 random forest training uses the asymmetry features, the seven posterior probability
tissue images output by the GMM
along with five additional geometric features based on the connected 
components of the segmentation image (distance image to tumor core,
physical volume image, volume to surface area ratio, eccentricity,
and elongation).
During the training phase (or RF model construction),
these feature images, $\left\{F_i: i \in GMM\right\}$, generated during Stage 1 are used to create
the Stage 1 RF model using the following relationship
\begin{align}
\label{eq:gmm}
 v_L \sim \sum_{i \in Asym} F_i + \sum_{i \in GMM} F_i + \sum_{i \in Misc} F_i
\end{align}
given in the standard \textit{R} notation of \cite{wilkinson1973}.  
In other words, $\bm{F}= [ \bm{F}_{Asym}, \bm{F}_{GMM},
\bm{F}_{Misc} ]$, where we column-wise concatenate the GMM, Asym and
Misc feature matrices.
The set of $\left\{F_i: i \in Asym\right\}$ are the feature images 
generated from the symmetric multivariate template 
described in the previous section and $\left\{F_i: i \in Misc\right\}$
are miscellaneous feature images described in Section \ref{sec:misc}.
Therefore, in total, the number of features is $k=FIXME$ where there
are 7 GMM features, FIXME asymmetry features and FIXME miscellaneous
geometric features.

\paragraph{Stage 2 (Refined spatial coherence classification)}
The first step in stage 2 is to use the per-class probabilities output
by the stage 1 random forest model as input to the Atropos
segmentation algorithm.  Application of the Stage 1 RF model to a set of feature images 
from an individual subject results in a set of $n$ spatial images
containing the voxel-wise probabilities for each 
of the seven tissue types.  At each voxel, the corresponding multi-feature 
vector is propagated through each tree of the random forest 
resulting in a single vote/classification per tree.  The
classification probabilities are derived by normalizing 
these votes to $[0,1]$ as in \cite{liaw2002}.

These spatial probability maps are passed as spatial priors to the
{\tt antsAtroposN4.sh} script in ANTs which couples the Atropos 
segmentation tool \citep{avants2011} with N4 bias correction \citep{tustison2010}.
The posterior probabilities that are output by Atropos and the
connected-component feature images for each modality (described in
section~\ref{sec:misc}) are then passed to the second stage random
forest model, along with the same asymmetry feature set based on our
symmetric templates.  
The second stage RF model is then produced similarly as the previous stage: 
\begin{align}
\label{eq:mapmrf}
\mathrm{Tissue}\,\,\mathrm{classes} \sim \sum_{i \in GMM_2} F_i + \sum_{i \in Asym} F_i + \sum_{i \in Misc} F_i,
\end{align}
where $GMM_2$ represents the stage 2 application of Atropos and the
other features are similar to those used in stage 1.
The use of these MRF features improves the accuracy of the 
intensity-based modeling over the GMM approach by introducing
a spatial coherence constraint to those features in addition to 
providing a better initialization with the RF posteriors from 
the initial estimation stage.  \textcolor{red}{This last sentence
  confuses me --- what is the difference between the GMM and MRF
  features?  I made a notation change $GMM_2$, b.c i thought the MRF
  and GMM were the same ... but maybe not?  did you exclude the MRF
  prior in the first application of Atropos?}

\subsubsection{Miscellaneous Multimodal Feature Image Generation}
\label{sec:misc}
Our strategy for feature selection was to, first, generate an ample feature
set and, second, to use the training phase to prioritize the features
appropriately.  In principal, this will eliminate weak features and
retain strong ones even in the presence of substantial noise.  Given
that image feature shape may be a valuable predictor of the presence of a
tumor \cite{zacharaki2009classification}, we calculate, for each modality, several images of neighborhood 
first-order statistics including mean, variance, skewness, and
entropy.  Neighborhoods are defined by block radii
of 1 and 3 voxels \textcolor{red}{Why these values?}.  We also calculated two Euclidean distance
transforms \citep{maurer2003} to be used for each subject.  One was
calculated from the subject's own cerebral mask and the other was the 
distance transform of the symmetric template cerebral mask warped
to the subject.  Finally, we also generated the  (T1 - T1C) difference 
image \citep{prastawa2003}.
The script {\tt createFeatureImages.sh} performs all preprocessing
and generates all feature images.  In Figure \ref{fig:featureImages} we provide sample mid-axial 
image slices from features generated from the {\tt BRATS\_HG0301} data
set.  Note that these feature images do not require the presence of
any training data.

@@

\begin{enumerate}
  \item Calculate stage 1 features based on:  
  \begin{itemize}
    \item GMM intensity modeling,
    \item asymmetry, and
    \item miscellaneous.
  \end{itemize}
  \item Apply stage 1 RF model to the above calculated features resulting in seven 
        probability maps for the seven tissue types.  
  \item Use the probability maps to calculate stage 2 features based on MAP-MRF intensity modeling.
  \item Apply stage 2 RF model to MAP-MRF, asymmetry, and miscellaneous features to 
  get final RF classification. 
\end{enumerate}

>>>>

\begin{figure}
  \centering
   \vspace{-15mm}
  \makebox[0.9\textwidth][c]{
    \includegraphics[width=150mm]{pipelineTestingTraining.pdf}
    }
  \caption{Diagrammatic workflow for the proposed RF model training (left) and prediction
  (right).  For training, the feature images are first generated from the set of input 
  training multimodal MRI and symmetric multivariate template. The set of features images are 
  use to create the RF model for the first stage.  This first stage RF model is then applied
  to the set of input training data to yield spatial priors for seeding the generation of the
  second stage feature images.  These feature images are then used to create the second
  stage RF model.  For prediction of a single-subject segmentation, a similar scenario 
  is applied whereby the generated feature images are used as input to the RF models the
  second of which produces an initial segmentation estimate.  This estimate is then refined
  using a series of heuristically-derived binary morphological operations resulting in the 
  final labeled brain.
  }
  \label{fig:pipeline}
\end{figure}

We use random forests as supervised learners.  Given ground truth,
voxel-wise $n$-class labels for a specific image, we can vectorize the
label image to produce $v_L$, the $p$-length vector of class labels
where $p$ is the number of voxels in the given brain of interest.
We also assume a matrix of features for the given brain, which we
denote $\bm{F}$ where $\bm{F}$ has dimensionality of $p$ by
$k$, where each of the $k$ columns corresponds a potentially valuable
feature vector.  Given this representation, the random forest model
can be trained, in \textit{R} notation, to predict the labels via
$v_L \sim  \bm{F}$.
Our segmentation protocol involves training and application
of two RF models in succession 
(see Figure \ref{fig:pipeline}) to which we refer as ``stages.''  The basic idea is that
we generate a set of feature images used as input to the
first RF model (or first stage) which produces a voxelwise 
probabilistic tissue estimate.  More precisely, in the RF framework, each prediction sample (i.e. the feature vector at each voxel),
is propagated through each tree of the ensemble where it is labeled as belonging to a specific
class.  
These ``votes'' are converted to voxelwise probabilistic
estimates for each class via standard mechanisms used in random forest
models, e.g. \cite{liaw2002}.
The seven labels of brain/tumor tissue type denote:
\begin{itemize}
\item cerebrospinal fluid,
\item grey matter,
\item white matter,
\item edema, 
\item non-enhancing tumor (including low-grade tumor center), 
\item enhancing tumor (excluding necrotic center), and 
\item abnormal necrotic center or necrocyst in high-grade gliomas.
\end{itemize}
We then use these output tissue map estimates as spatial 
priors for generating a second set of geometrically refined image features.  These
are used as input for a second RF model application (or
second stage),
the output of which constitute the final tumor segmentation estimate.
Note that some feature images are included directly and
  without modification in both stages such as
the asymmetry features described in the previous section.

@@

\paragraph{Stage 1:  Voxelwise classification with Gaussian mixture
  modeling} 
The first step in processing a new image is to segment the image into
the $n$ tissue classes listed above.   We achieve this via an expectation-maximization
segmentation method based on
Gaussian mixture modeling (GMM).  GMM results in voxelwise
tissue probabilities which comprise an additional set of image
features.  The GMM is initialized with prior cluster centers for
specific tissue types.  We learn the values for these cluster centers
for each modality and each tissue from training data
\citep{reynolds2009}.  We also perform intensity normalization
across the training cohort by winsorizing the intensity values to the quantile
range $[0.01, 0.99]$ and then rescaling the resulting intensities to 
$[0, 1]$.  The cluster centers are defined as the mean
normalized intensity value for each tissue type of each modality 
image over all the training data.  We tried a more sophisticated intensity
normalization scheme using the approach of \cite{nyul2000} but found that
our simpler normalization approach, in combination with the rest of the
pipeline, produced slightly better results.

More formally, the GMM computes the 
probability distribution at each voxel, $\mathbf{x}$, as the
sum of $M$ Gaussian components, $\mathcal{N}(\mathbf{x}|\mu,\sigma)$, i.e.
\begin{align}
p\left(\mathbf{x}|\mu_m,\sigma_m,\lambda_m\right) = \sum_{i=1}^M \lambda_m \mathcal{N}(\mathbf{x}|\mu_m,\sigma_m)
\end{align}
where $\sum_{m=1}^M \lambda_m = 1$.  The parameters of the GMM 
are determined using the Atropos segmentation tool \citep{avants2011} available in ANTs.
For the BRATS data, GMM modeling was applied to each of the four modalities, i.e.,
FLAIR, T1, T1 contrast, and T2 (see Figure \ref{fig:stageComparison}), separately.  
Note that no spatial priors are used during this stage including MRF spatial priors.

%\paragraph{Stage 1 random forest} 
The stage 1 random forest training uses the asymmetry features, the seven posterior probability
tissue images output by the GMM
along with five additional geometric features based on the connected 
components of the segmentation image (distance image to tumor core,
physical volume image, volume to surface area ratio, eccentricity,
and elongation).
During the training phase (or RF model construction),
these feature images, $\left\{F_i: i \in GMM\right\}$, generated during Stage 1 are used to create
the Stage 1 RF model using the following relationship
\begin{align}
\label{eq:gmm}
 v_L \sim \sum_{i \in Asym} F_i + \sum_{i \in GMM} F_i + \sum_{i \in Misc} F_i
\end{align}
given in the standard \textit{R} notation of \cite{wilkinson1973}.  
In other words, $\bm{F}= [ \bm{F}_{Asym}, \bm{F}_{GMM},
\bm{F}_{Misc} ]$, where we column-wise concatenate the GMM, Asym and
Misc feature matrices.
The set of $\left\{F_i: i \in Asym\right\}$ are the feature images 
generated from the symmetric multivariate template 
described in the previous section and $\left\{F_i: i \in Misc\right\}$
are miscellaneous feature images described in Section \ref{sec:misc}.
Therefore, in total, the number of features is $k=84$ where there
are twelve GMM features for each of the four modalities ($n=48$), 
two asymmetry features for each modality ($n=8$), and six modality-specific 
miscellaneous geometric features ($n=24$) with an additional four feature
images.

% 48 + 8 + 24 + 4 = 84

\paragraph{Stage 2:  Refined spatial coherence classification}
The first step in stage 2 is to use the per-class probabilities output
by the stage 1 random forest model as input to the Atropos
segmentation algorithm.  Application of the Stage 1 RF model to a set of feature images 
from an individual subject results in a set of $n$ spatial images
containing the voxel-wise probabilities for each 
of the seven tissue types.  At each voxel, the corresponding multi-feature 
vector is propagated through each tree of the random forest 
resulting in a single vote/classification per tree.  The
classification probabilities are derived by normalizing 
these votes to $[0,1]$ as in \cite{liaw2002}.

These spatial probability maps are passed as spatial priors to the
{\tt antsAtroposN4.sh} script in ANTs which couples the Atropos 
segmentation tool \citep{avants2011} with N4 bias correction \citep{tustison2010}.
The posterior probabilities that are output by Atropos and the
connected-component feature images for each modality (described in
section~\ref{sec:misc}) are then passed to the second stage random
forest model, along with the same asymmetry feature set based on our
symmetric templates.  
The second stage RF model is then produced similarly as the previous stage: 
\begin{align}
\label{eq:mapmrf}
\mathrm{Tissue}\,\,\mathrm{classes} \sim \sum_{i \in MAP\_MRF} F_i + \sum_{i \in Asym} F_i + \sum_{i \in Misc} F_i,
\end{align}
where $MAP-MRF$ represents the Stage 2 application of Atropos and the
other features are similar to those used in Stage 1.
The use of these MRF features improves the accuracy of the 
intensity-based modeling over the GMM approach by introducing
a spatial coherence constraint (the MRF prior) to those features 
in addition to 
providing a better initialization with the RF posteriors from 
the initial estimation stage (which lacked the MRF spatial prior).

\subsubsection{Miscellaneous Multimodal Feature Image Generation}
\label{sec:misc}
Our strategy for feature selection was to, first, generate an ample feature
set and, second, to use the training phase to prioritize the features
appropriately.  In principal, this will eliminate weak features and
retain strong ones even in the presence of substantial noise.  Given
that image feature shape may be a valuable predictor of the presence of a
tumor \cite{zacharaki2009classification}, we calculate, for each modality, several images of neighborhood 
first-order statistics including mean, variance, skewness, and
entropy.  Block neighborhood radii of values $\{1,2,3\}$ were tested.
Due to overlapping discriminative qualities, we narrowed the feature
values to 1 and 3 voxels.  We also calculated two Euclidean distance
transforms \citep{maurer2003} to be used for each subject.  One was
calculated from the subject's own cerebral mask and the other was the 
distance transform of the symmetric template cerebral mask warped
to the subject.  Finally, we also generated the  (T1 - T1C) difference 
image \citep{prastawa2003}.
The script {\tt createFeatureImages.sh} performs all preprocessing
and generates all feature images.  In Figure \ref{fig:featureImages} we provide sample mid-axial 
image slices from features generated from the {\tt BRATS\_HG0301} data
set.  Note that these feature images do not require the presence of
any training data.

@@

\begin{enumerate}
  \item Calculate Stage 1 features based on:  
  \begin{itemize}
    \item GMM intensity modeling,
    \item asymmetry, and
    \item miscellaneous.
  \end{itemize}
  \item Apply Stage 1 RF model to the above calculated features resulting in seven 
        probability maps for the seven tissue types.  
  \item Use the probability maps to calculate Stage 2 features based on MAP-MRF intensity modeling.
  \item Apply Stage 2 RF model to MAP-MRF, asymmetry, and miscellaneous features to 
  get final RF classification. 
\end{enumerate}
